\Large \bf Vigor: Toward Carrier-Grade Software Network Functions

### 2016-11-24 doc/paper/paper.tex
Another round of Luis's feedback for the Solution Overview section.

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{listings}
\usepackage[]{algorithm2e}
\usepackage{changepage}
\usepackage{hyperref}
\definecolor{mygreen}{rgb}{0.1,0.5,0.1}
\definecolor{myblue}{rgb}{0.1,0.1,0.7}

@@

  while(1) {^\label{line:loop-begin}^
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) && ^\label{line:recv-pkt}^
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len && can_send_packet()) { ^\label{line:can-send}^
      send_packet(&packets[begin]); ^\label{line:send-pkt}^
      begin = (begin + 1)%CAP;
    }
  }^\label{line:loop-end}^
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop (l.\ref{line:loop-begin}-\ref{line:loop-end}) and a latent ring
    implementation (declared on l.\ref{line:pkt-def},\ref{line:iter-def})}
  \label{lst:orig}
\end{figure}

In order to absorb bursts, the code employs a ring buffer. The application
interacts with the network using the following 3 functions:
\code{receive_packet}(l.\ref{line:recv-pkt}), which non-blockingly reads an
inbound packet and stores it at \code{packets[end]}, returning success or
failure; \code{can_send_packet}(l.\ref{line:can-send}), which checks if a new
packet can be sent; and \code{send_packet}(l.\ref{line:send-pkt}), which sends
the packet stored at \code{packets[begin]}. The application follows a common
practice: it consists of a main event loop, it has no other complex loops, and
it uses a ring, a popular data structure in network programming.

Our NF accumulates state in the ring thus making many existing verification
approaches impractical.

Fig.~\ref{lst:vigor} shows the actual code that we are going to verify. The
first two steps show how we get there. \code{VIGOR_CHECK}
(l.~\ref{line:vigcheck}) adds additional property that we would like to verify
--- that we do not forward packets with port 9.

\begin{figure}[h!]
\begin{lstlisting}
int main(int argc, char** argv) {
  struct packet p;
  struct ring *r = ring_create(CAP);^\label{line:mk-ring}^
  if (!r) return 1;
  while(LOOP(1))^\label{line:loop-hd}^
  {
    loop_iteration_begin(&r);
    if (!ring_full(r))
      if (receive_packet(&p) && p.port != 9)^\label{line:if-cond}^
        ring_push_back(r, &p);
    if (!ring_empty(r) && can_send_packet()) {
      ring_pop_front(r, &p);^\label{line:pop-front}^
      VIGOR_CHECK(p.port != 9);^\label{line:vigcheck}^
      send_packet(&p);
    }
    loop_iteration_end(&r);
  }
  return 0;
}
\end{lstlisting}
  \caption{Vigor implementation of Drop protocol.
    \code{LOOP}(l.\ref{line:loop-hd}) labels the infinite loop head and we
    explicitly use a ring(l.\ref{line:mk-ring}) from the Vigor library. \code{VIGOR_CHECK} at the
    line \ref{line:vigcheck} adds a semantic property to be verified.}
  \label{lst:vigor}
\end{figure}

{\bf Step 1: Label infinite loops.} Our approach requires special handling of
infinite loops in the application code. A developer puts a label \code{LOOP}
(l.\label{line:loophd}) into the condition in order to designate the
potentially infinite loop. For a typical NF there should be only a handful or
them (1 in our example).

@@

static bool^\label{line:c-user-prop-from}^
packet_constraints(struct packet* p) {
  return p->port != 9;^\label{line:c-pkt-cnstr}^
}^\label{line:c-user-prop-to}^
\end{lstlisting}
  \caption{Custom definitions capturing the desired property. They capture the
    essence of every packet in the ring. Currently, we have to duplicate the
    packet condition in two languages (formal logic --
    ll\ref{line:vf-user-prop-from}-\ref{line:vf-user-prop-to}, and C --
    ll\ref{line:c-user-prop-from}-\ref{line:c-user-prop-to}) due to the
    limitations of our prototype.}
  \label{lst:customizations}
\end{figure}

@@

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. Another classic SE challenge ---
``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset} spoils
the party here. When indexing an array with a symbolic variable (or when
dereferencing a symbolic pointer), the constraints on program state become
extremely complex, because the symbolic pointer or index could be referencing
many different possible locations. The complexity of the constraints chokes the
SMT solver, and the SEE grinds to a halt. In general, SE is not practical for
code with symbolic pointers or offsets. Alas, stateful C code almost always has
some of these.

In this example the symbolic indexing comes from the ring array. Once we
replaced the infinite loop with a single iteration we have to assume (due to
havoc-ing) the \code{begin} and \code{begin + queue_len} indexes may point
anywhere in the array.

In Vigor, we circumvent the problem of symbolic indexing by encapsulating program
state into well-defined data structures and putting them into a ``library of
components'' (step 2). This works well for NF development, because the NF code
uses a relatively small set of standard structures, such as ring buffers, hash
tables, least-prefix-match tables, etc. We provide to the SEE symbolic models of
these abstracted components that ``hide'' symbolic indexes and pointers from the
SEE.

After the step 2 exhaustive SE takes less than 100 millisec regardless of the
value of \code{CAP}. The fact that the code does not
crash on any path amounts to a proof of crash freedom. Or does it? The validity
of the proof depends on our library implementation of \code{ring} not
crashing; and (2) the symbolic model of \code{ring} accurately capturing how the
\code{ring} implementation works.

@@

Here is an example of the proven specification for the function
\code{ring_pop_front}:

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p)
/*@ requires ringp(r, ?user_property,^\label{line:precondition-from}^
                   ?lst, ?cap) &*&
             lst != nil &*& packetp(p, _); @*/^\label{line:precondition-to}^
/*@ ensures ringp(r, user_property,^\label{line:postcondition-from}^
                  tail(lst), cap) &*&
            packetp(p, head(lst)) &*&
            true == user_property(head(lst)); @*/^\label{line:postcondition-to}^
{
  //@ extract_first(r); ^\label{line:lemma1}^
  struct packet* src_pkt = r->array + r->begin;
  p->port = src_pkt->port;
  r->len = r->len - 1;
  r->begin = r->begin + 1;
  if (r->cap <= r->begin) {
    r->begin = 0;
    //@ stitch_with_empty_overflow(r);^\label{line:lemma2}^
  } else {
    //@ stitch_with_empty(r);^\label{line:lemma3}^
  }
}
\end{lstlisting}
  \caption{A formal contract and verified implementation of the
    \code{ring_pop_front} function. The comments starting with \code{@} symbol
    are formal annotations that hold the function contract
    (\code{requires}(ll.\ref{line:precondition-from}-\ref{line:precondition-to})
    - precondition,
    \code{ensures}(ll.\ref{line:postcondition-from}-\ref{line:postcondition-to})
    - postcondition) and
    lemmas(ll.\ref{line:lemma1},\ref{line:lemma2},\ref{line:lemma3}) that help
    the proof checker to verify the state transformations in the code.}
  \label{lst:contract}
\end{figure}

This C code is intermingled with the correctness proof annotations that help the
formal verification engine (VeriFast) with small steps to transform the initial
state, given in \code{requires} to the final state, specified in \code{ensures}.
After proving the correctness of this transformation, the formal verification
engine will use \code{requires/ensures} formulas as pre- and post- conditions
for \code{ring_pop_front} contract. The trigraph \code{&*&} separates different
statements that hold together. For example

\begin{lstlisting}
ringp(r, user_property, tail(lst), cap) &*&
packetp(p, head(lst)) &*&
true == user_property(head(lst));
\end{lstlisting}
means:
\begin{itemize}
\item \code{r} points to a healthy ring of capacity \code{cap}, that contains
  \code{tail(lst)} packets, that satisfy the \code{user_property}.
\item \code{p} points to a packet ``\code{head(lst)}''.
\item ``\code{head(lst)}'' satisfies the \code{user_property}.
\end{itemize}

@@

You can see one of the generated call sequences on Fig.~\ref{lst:call-sequence}.
It starts with \code{loop_invariant_produce} that marks the beginning of a loop
iteration. It continues then with calls to \code{ring_full} and \code{ring_empty},
noting (with \code{assume}) that both return \code{false}. The seqeunce ends
with a call to \code{ring_pop_front}. The last line detects underapproximation
of the model output.

@@

loop_invariant_produce(&(arg1));
/*@ open loop_invariant(_); @*/
bool ret1 = ring_full(arg1);
//@ assume(ret1 == false);
bool ret2 = ring_empty(arg1);
//@ assume(ret2 == false);
/*@ close packetp(&(arg2),
                  packet((&(arg2))->port)); @*/

ring_pop_front(arg1, &(arg2));
//@ open packetp(&(arg2), _);

/*@ assert(arg2.port != 9); @*/^\label{line:assert}^
\end{lstlisting}
  \caption{A call sequence, produced by SEE. Vigor adds formal annotations to
    constrain the execution space.}
  \label{lst:call-sequence}
\end{figure}

@@

Let us follow now the `pord != 9' condition across the framework. During
exhaustive SE, SEE first learns the constraint in the \code{if} condition:
Fig.~\ref{lst:vigor} l.\ref{line:if-cond}. It however forgets it on the next
line, because our model of \code{ring_push_back} discards the packet being
pushed. However, this constraint arises again on the call to
\code{ring_pop_front}(l.\ref{line:pop-front}) as
\code{packet_constraints}(Fig.~\ref{lst:customizations},l.\ref{line:c-pkt-cnstr}),
according to the symbolic model - Fig.~\ref{lst:symbolic-model}. Being applied
to the fresh packet from the ring, it satisfies the \code{VIGOR_CHECK}
(Fig.~\ref{lst:vigor} l.\ref{line:vigcheck}).

@@

\subsection{Misfit symptoms}

Consider potential model mismatches. Suppose we had an overapproximate model ---
Fig.~\ref{lst:overapproximate}. Then the \code{VIGOR_CHECK}
(Fig.\ref{lst:vigor}) in the application code would fail.
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  // No constraints on the packet port here.
}
\end{lstlisting}
  \caption{An overapproximate ring model fragment. It does not add the
    constraint necessary for the code to verify.}
  \label{lst:overapproximate}
\end{figure}

@@

Suppose now we had an underapproximate model: Fig.~\ref{lst:underapproximate}.
Then the last line in the call sequence check (Fig.\ref{lst:call-sequence}) would look like:
\begin{lstlisting}
/*@ assert(arg2.port == 0); @*/
\end{lstlisting}
The assertion fails, because the contract (Fig.~\ref{lst:contract}) specifies
a wider range for \code{arg2.port} than just \code{0}.

>>>>

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix,epsfig}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{listings}
\usepackage[]{algorithm2e}
\usepackage{changepage}
\usepackage{hyperref}
\usepackage[all]{hypcap}        % needed to help hyperlinks direct correctly
\definecolor{mygreen}{rgb}{0.1,0.5,0.1}
\definecolor{myblue}{rgb}{0.1,0.1,0.7}

@@

  while(1) {^\label{line:loop-begin}^
    if (queue_len < CAP) {
      int end = (begin + queue_len)%CAP;
      if (receive_packet(&packets[end]) && ^\label{line:recv-pkt}^
          packets[end].port != 9)
        queue_len += 1;
    }
    if (0 < queue_len && can_send_packet()) { ^\label{line:can-send}^
      send_packet(&packets[begin]); ^\label{line:send-pkt}^
      begin = (begin + 1)%CAP;
    }
  }^\label{line:loop-end}^
  return 0;
}
\end{lstlisting}
  \caption{Original implementation of the Drop protocol. Contains an infinite
    loop (ll.\ref{line:loop-begin}-\ref{line:loop-end}) and a latent ring
    implementation (declared on l.\ref{line:pkt-def},\ref{line:iter-def})}
  \label{lst:orig}
\end{figure}

In order to absorb bursts, the code employs a ring buffer. The application
interacts with the network using the following 3 functions:
\code{receive_packet} (l.\ref{line:recv-pkt}), which non-blockingly reads an
inbound packet and stores it at \code{packets[end]} -- the end of the ring
buffer, returning success or failure; \code{can_send_packet}
(l.\ref{line:can-send}), which checks if a new packet can be sent; and
\code{send_packet} (l.\ref{line:send-pkt}), which sends the packet stored at
\code{packets[begin]} -- the beginning of the ring buffer. The application
follows a common practice: it consists of a main event loop, it has no other
complex loops, and it uses a ring, a popular data structure in network
programming. As a result, our NF accumulates state in the ring thus making many
existing verification approaches impractical.

Fig.~\ref{lst:vigor} shows the annotated code, ready for the verification with
Vigor. It is the result of applying the first two steps shown below.
\code{VIGOR_CHECK} (l.\ref{line:vigcheck}) adds additional property that we
would like to verify --- that we do not forward packets with port 9.

\begin{figure}[h!]
\begin{lstlisting}
int main(int argc, char** argv) {
  struct packet p;
  struct ring *r = ring_create(CAP);^\label{line:mk-ring}^
  if (!r) return 1;
  while(LOOP(1))^\label{line:loop-hd}^
  {
    loop_iteration_begin(&r);
    if (!ring_full(r))
      if (receive_packet(&p) && p.port != 9)^\label{line:if-cond}^
        ring_push_back(r, &p);
    if (!ring_empty(r) && can_send_packet()) {
      ring_pop_front(r, &p);^\label{line:pop-front}^
      VIGOR_CHECK(p.port != 9);^\label{line:vigcheck}^
      send_packet(&p);
    }
    loop_iteration_end(&r);
  }
  return 0;
}
\end{lstlisting}
  \caption{Vigor implementation of Drop protocol.
    \code{LOOP} (l.\ref{line:loop-hd}) labels the infinite loop head and we
    explicitly use a ring (l.\ref{line:mk-ring}) from the Vigor library. \code{VIGOR_CHECK} at the
    line \ref{line:vigcheck} adds a semantic property to be verified.}
  \label{lst:vigor}
\end{figure}

{\bf Step 1: Label infinite loops.} Our approach requires special handling of
infinite loops in the application code. A developer puts a label \code{LOOP}
(l.\ref{line:loop-hd}) into the condition in order to designate the
potentially infinite loop. For a typical NF there should be only a handful or
them (1 in our example).

@@

static bool^\label{line:c-user-prop-from}^
packet_constraints(struct packet* p) {
  return p->port != 9;^\label{line:c-pkt-cnstr}^
}^\label{line:c-user-prop-to}^
\end{lstlisting}
  \caption{Custom definitions capturing the desired property. They capture the
    essence of every packet in the ring. Currently, we have to duplicate the
    packet condition in two languages (formal logic --
    ll.\ref{line:vf-user-prop-from}-\ref{line:vf-user-prop-to}, and C --
    ll.\ref{line:c-user-prop-from}-\ref{line:c-user-prop-to}) due to the
    limitations of our prototype.}
  \label{lst:customizations}
\end{figure}

@@

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. Another classic SE challenge ---
``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset} spoils
the party here. When indexing an array with a symbolic variable (or when
dereferencing a symbolic pointer), the constraints on program state become
extremely complex, because the symbolic pointer or index could be referencing
many different possible locations. The complexity of the constraints chokes the
SMT solver, and the SEE grinds to a halt. In general, SE is not practical for
code with symbolic pointers or offsets. Alas, stateful C code almost always has
some of these. In this example the symbolic indexing comes from the ring array.
Once we replaced the infinite loop with a single iteration we have to assume
(due to havoc-ing) the \code{begin} and \code{begin + queue_len} indexes may
point anywhere in the array.

In Vigor, we circumvent the problem of symbolic indexing by encapsulating
program state into well-defined data structures and putting them into a
``library of components'' (e.g. the ring in the step 2). This works well for NF
development, because the NF code uses a relatively small set of standard
structures, such as ring buffers, hash tables, least-prefix-match tables, etc.
We provide to the SEE symbolic models of these abstracted components that
``hide'' symbolic indexes and pointers from the SEE.

After the step 2 exhaustive SE takes less than 100 millisec regardless of the
value of \code{CAP}. The fact that the code does not crash on any path amounts
to a proof of crash freedom. Or does it? The validity of the proof depends on
our library implementation of \code{ring} not crashing; and the symbolic model
of \code{ring} accurately capturing how the \code{ring} implementation works.

@@

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p)
/*@ requires ringp(r, ?user_property,^\label{line:precondition-from}^
                   ?lst, ?cap) &*&
             lst != nil &*& packetp(p, _); @*/^\label{line:precondition-to}^
/*@ ensures ringp(r, user_property,^\label{line:postcondition-from}^
                  tail(lst), cap) &*&
            packetp(p, head(lst)) &*&
            true == user_property(head(lst)); @*/^\label{line:postcondition-to}^
{
  //@ extract_first(r); ^\label{line:lemma1}^
  struct packet* src_pkt = r->array + r->begin;
  p->port = src_pkt->port;
  r->len = r->len - 1;
  r->begin = r->begin + 1;
  if (r->cap <= r->begin) {
    r->begin = 0;
    //@ stitch_with_empty_overflow(r);^\label{line:lemma2}^
  } else {
    //@ stitch_with_empty(r);^\label{line:lemma3}^
  }
}
\end{lstlisting}
  \caption{A formal contract and verified implementation of the
    \code{ring_pop_front} function. The comments starting with \code{@} symbol
    are formal annotations that hold the function contract
    (\code{requires} (ll.\ref{line:precondition-from}-\ref{line:precondition-to})
    - precondition,
    \code{ensures} (ll.\ref{line:postcondition-from}-\ref{line:postcondition-to})
    - postcondition) and
    lemmas (ll.\ref{line:lemma1},\ref{line:lemma2},\ref{line:lemma3}) that help
    the proof checker to verify the state transformations in the code.}
  \label{lst:contract}
\end{figure}

Figure~\ref{lst:contract} shows the formal specification and a proven
implementation of the example function \code{ring_pop_front}. This C code is
intermingled with the correctness proof annotations that help the formal
verification engine (VeriFast) with small steps to transform the initial state,
given in \code{requires} to the final state, specified in \code{ensures}. After
proving the correctness of this transformation, the formal verification engine
will use \code{requires/ensures} formulas as pre- and post- conditions for
\code{ring_pop_front} contract. The trigraph \code{&*&} separates different
statements that hold together. For example

\begin{lstlisting}
ringp(r, user_property, tail(lst), cap) &*&
packetp(p, head(lst)) &*&
true == user_property(head(lst));
\end{lstlisting}
means:
\begin{itemize}
\item \code{r} points to a healthy ring of capacity \code{cap}, that contains
  \code{tail(lst)} packets, that all satisfy the \code{user_property}.
\item \code{p} points to a packet ``\code{head(lst)}''.
\item ``\code{head(lst)}'' satisfies the \code{user_property}.
\end{itemize}

@@

You can see one of the generated call sequences on Fig.~\ref{lst:call-sequence}.
It starts with \code{loop_invariant_produce} that is inserted automatically at
the beginning of a loop iteration. It continues then with calls to
\code{ring_full} and \code{ring_empty}, noting (with \code{assume}) that both
return \code{false}. They both return false in this particular trace, in other
traces they returned other values, but to define the context for the last
function call (l.\ref{line:tip-call}) we must repeat the history as closely as
possible. Vigor handles the result of the last call in the sequence very
differently. Instead of just telling the formal verification engine what
happened, it inserts an \code{assert}(l.\ref{line:assert}) reflecting the
symbolic model output to check that it is consistent with the postcondition of
the last function (Fig.~\ref{lst:contract}).

@@

loop_invariant_produce(&(arg1));
/*@ open loop_invariant(_); @*/^\label{line:unexplained1}^
bool ret1 = ring_full(arg1);
//@ assume(ret1 == false);
bool ret2 = ring_empty(arg1);
//@ assume(ret2 == false);
/*@ close packetp(&(arg2),^\label{line:unexplained2}^
                  packet((&(arg2))->port)); @*/^\label{line:unexplained3}^

ring_pop_front(arg1, &(arg2));^\label{line:tip-call}^
//@ open packetp(&(arg2), _);^\label{line:unexplained4}^

/*@ assert(arg2.port != 9); @*/^\label{line:assert}^
\end{lstlisting}
  \caption{A call sequence, produced by SEE. Vigor adds formal annotations to
    constrain the execution space. Unexplained
    lines(\ref{line:unexplained1},\ref{line:unexplained2},\ref{line:unexplained3},\ref{line:unexplained4})
    are the automatically inserted helper steps to guide the proof checker.}
  \label{lst:call-sequence}
\end{figure}

@@

Let us follow now the `port != 9' condition across the framework. During
exhaustive SE, SEE first learns the constraint in the \code{if} condition:
Fig.~\ref{lst:vigor} l.\ref{line:if-cond}. Here, SEE just records it in the call
sequence for model validation, and does not use further in the SE, because our
model of \code{ring_push_back} discards the packet being pushed. However, the
call to \code{ring_pop_front} (l.\ref{line:pop-front}) reintroduces this
constraint formulated in \code{packet_constraints}
(Fig.~\ref{lst:customizations},l.\ref{line:c-pkt-cnstr}), according to the
symbolic model - Fig.~\ref{lst:symbolic-model}. Being applied to the fresh
packet from the ring, it satisfies the \code{VIGOR_CHECK} (Fig.~\ref{lst:vigor}
l.\ref{line:vigcheck}).

@@

\subsection{Model inconsystency symptoms}

The component developer produces a number of model that capture the component
semantics in more or less details for different usecases. For a particular use
case --- a NF, a particular model may be inappropriate. An inappropriate model
may not compromise the verification result, because Vigor detects both
overapproximate and underapproximate models.

Suppose we had an overapproximate model --- Fig.~\ref{lst:overapproximate}. Then
the popped packet would come unconstrained and the \code{VIGOR_CHECK}
(Fig.\ref{lst:vigor} l.\ref{line:vigcheck}) in the application code would fail.
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  // No constraints on the packet port here.
}
\end{lstlisting}
  \caption{An overapproximate ring model fragment. It does not add the
    constraint necessary for the code to verify.}
  \label{lst:overapproximate}
\end{figure}

@@

Suppose now we had an underapproximate model: Fig.~\ref{lst:underapproximate}.
Then the last line in the call sequence check (Fig.\ref{lst:call-sequence}) in
this case would look different:
\begin{lstlisting}
/*@ assert(arg2.port == 0); @*/
\end{lstlisting}
The assertion fails, because the contract (Fig.~\ref{lst:contract}) specifies
a wider range for \code{arg2.port} than just \code{0}.

Different models may work well for different usecases. E.g. the overapproximate
model is just right for a buffer, that does not drop the packets with port 9.
The underapproximate model would be just right for a buffer that forwards only the
packets with port equeal to 0.
